# Propositions to frame uncertainty {#propositions}

This Chapter introduces propositions

## Facts-in-issue

To motivate the use of propositions, we begin with a discussion about facts-in-issue. Facts-in-issue are the relevant facts of a specific case which are contested by each side of the legal dispute. In criminal trials, this typically manifests as the defence contesting facts which the prosecution have asserted.

There is epistemic uncertainty about the facts-in-issue and so relevant evidence is presented to the court. The fact finder then considers this evidence to determine the truth of the facts-in-issue, and ultimately the outcome of the case. 

It is clear from this that the language and presentation of facts-in-issue must be precise and accurate. The presentation should also be done in a way that allows the fact finder to clearly discriminate between what either side of counsel is asserting. To achieve this, forensic scientists formalise the facts-in-issue into what are known as propositions. 

## Propositions

A proposition is an assertion about a factual state of nature. The 'factual' part of that is key since it means that the assertion is capable of being either true or false. Thus, propositions are capable of either being true or false. Although not always correct, we will speak of propositions as not only **capable of being** either true or false, but **as** being either true or false. 

The truth of propositions does not depend upon people's belief about them (unless the proposition is specifically about their belief). Consider the following:

1. The Earth orbits the Sun.
2. The Earth is flat.

Proposition 1 is true and proposition 2 is false, and we have substantial scientific evidence to prove this. At a certain point of time in history, however, common belief about those propositions was false: people thought that the Sun orbited the Earth, and that the Earth was flat (and some still do...). It was true that the Earth orbits the Sun both before and after humans had enough evidence to believe it was true. 

Propositions can still evoke epistemic uncertainty; they are true or false but we might not know which. In this case we can probabilistically reason about them. Consider another example:

3. Poker Player B is holding an ace in this round.

In this hypothetical poker game, it's self-evident that the proposition is true if Player B is holding an ace and it is false if Player B is not holding an ace. If you are Player B then you will know the truth of this and the proposition is trivial. However, if you are not Player B then you won't know the truth of this; you have epistemic uncertainty.

To play your hand well, you will have to come to a reasonable belief about the probability of this proposition (and others) and make decisions about checking, raising, or folding etc., whilst accounting for this uncertainty. Note that in this example you might **never** know the truth of this proposition. Player B might never turn over their cards nor reveal the answer. The proposition was still either true or false regardless of this.

The truth of some propositions may change as situations change. Consider the following:

4. Boris Johnson's Conservative party govern the UK. 

As of me writing this, the proposition is true. The UK still exists, the Conservative party exist under that same name and are led by Boris Johnson, and his party governs the UK after a landslide general election win. This proposition won't be true forever, and may not even be true by the time you're reading this. It can help to be more specific when formulating propositions to overcome this dependence on time or other factors. Changing the proposition instead to:

5. On 20th May 2020, Boris Johnson's Conservative party govern the UK 

gives a more specific proposition that is now unambigious about the temporal factor which may affect it. However, the proposition is now less general since it only refers to one specific day. Whether this lack of generality is worth the extra specificity or not depends on the question at hand. When the question is provided by investigators and given to forensic scientists, then the scientists have to balance these factors when formulating appropriate propositions. 

## Competing propositions

When formulating propositions from facts-in-issue, the different sides of the adversarial system make different assertions about the circumstances which lead to the same factual observation.

For example, as part of a case there might be a DNA sample at the crime scene which was found to match the DNA profile taken from the suspect after their arrest. The prosecution might assert that the defendant was at the location in which the DNA sample was recovered. The defence, on the other hand, might accept that this is the defendant's DNA but instead assert that it was indirectly transferred there, e.g. from contact with another individual who was provably at the crime scene. Both of these assertions would reasonably lead to the defendant's DNA matching the crime scene sample.

Translating these assertions into propositions gives so-called **competing propositions**. Competing propositions typically come in pairs, one put forward by each side of legal counsel. They are often denoted as $H_p$ and $H_d$ to represent 'prosecution/defence hypothesis', and we'll use this from now on.

Case scenario: Police were called to the house of a known local drug-dealer after reports by neighbours of a suspected gunshot heard from within. They arrived at the scene to find a middle-aged male, the known drug-dealer, shot dead in a bedroom. A window at the back of the house on the ground floor was found to be smashed, with the glass having fallen towards the inside of the house. They found open death-threats on the deceased's public social media account from two males. The males are known to police, having previously been involved in drug-related crimes too. Both are arrested at their known addresses, at which time Suspect 1 confesses to having broken in to the house but not having fired the weapon. He refuses to name any accomplices. Suspect 2 refuses to comment, despite glass-covered clothing being found within a plastic bag in a bin at his residence. After further testing, this clothing was also found to be covered in gunshot-residue particles. No weapon is recovered.

Here are three facts-in-issue of the (probably) many for this case at this preliminary stage. 

Fact-in-issue 1: whether or not Suspect 2 killed the victim with the intent to kill or cause grievous bodily harm.

- $H_p^1$: Suspect 2 committed murder,
- $H_d^1$: Suspect 2 did not commit murder,

Fact-in-issue 2: whether or not Suspect 2 fired the gun in question

- $H_p^2$: Suspect 2 fired the gun,
- $H_d^2$: Suspect 2 did not fire the gun,

Fact-in-issue 3: whether or not the glass fragments found on Suspect 2's clothing originated from the smashed window

- $H_p^3$: The glass fragments originate from the smashed window,
- $H_d^3$: The glass fragments originate from a source other than the smashed window.

We have extracted the information from the case scenario and formulated potential facts-in-issue. These facts-in-issue isolate individual issues that appear in the case scenario. They also make a clear statement about what is the potential issue which surrounds each fact, e.g. was it or was it not the suspect?

It is clear that for each fact-in-issue, we have created a potential pair of competing propositions. Each proposition makes a direct assertion about the associated fact-in-issue, considers each side of the issue (prosecution or defence) separately, and is either true or false. Using this process reduces complex case scenarios into individual unambiguous points of focus, around which the assessment of evidence can be anchored.

There are some other key properties of these propositions which are worth exploring. Let's revisit the following competing pair of propositions:

- $H_p^2$: Suspect 2 fired the gun,
- $H_d^2$: Suspect 2 did not fire the gun.

One property of these propositions is that when one is true, the other is necessarily false (and vice versa). They cannot both be true. This is known as **mutual exclusivity**. Mutual exclusivity occurs when two propositions make assertions which have no logical overlap. Suspect 2 firing the gun and not firing the gun cannot both be true, they are completely separable assertions.

Another property of these propositions is that either one or the other must be true -- there is no scenario in which both are false. This is known as **exhaustivity**. Exhaustivity means that the propositions cover the entire set of all possibilities for the event in question. We know that a gun was fired, so either Suspect 2 fired the gun or they did not. The proposition $H_p^2$ covers one specific event and $H_d^2$ covers everything else. 

When competing propositions are exhaustive and mutually exclusive then they cover all possible events, meaning that one of them must be true, but they cannot both be true. This means that one of the propositions must be true and the other must be false. This 'either one or the other' property makes such competing propositions easier to evaluate, since knowing the truth of one determines the truth of the other.

These propositions were guaranteed to be mutually exclusive and exhaustive because of the way we constructed them. First we specified $H_p$. Then, we **negated** $H_p$ to get $H_d$ in each pair of propositions. A simpler way of phrasing this is to say we formed '$H_p$' and 'not $H_p$'. Take the glass propositions as an example:

- $H_p^3$: The glass fragments originate from the smashed window,
- $H_d^3$: The glass fragments originate from a source other than the smashed window.

If $H_p^3$ is not true, then the glass fragments must originate from a source other than the smashed window, and so we get $H_d^3$. This makes the propositions necessarily mutually exclusive, '$H_p^3$' and 'not $H_p^3$' cannot both be true because they are logical negations. It also means that they are exhaustive, since $H_p^3$ or its negation ('anything else') must be true.

The logic of negating propositions is in line with the burden of proof in UK criminal cases. It is for the prosecution to prove beyond reasonable doubt that their assertions are true. The defence, in proving reasonable doubt about the prosecution assertions need only show that the opposite of the prosecution story has reasonable credibility. Negating prosecution propositions is one logical way of doing this.

There are plenty of situations in which simply negating the prosecution propositions is inappropriate however, e.g. when the defendant has a genuine alternative narrative for events. In this situation, the defence narrative might be used for $H_d$, e.g.:

- $H_d^{3^*}$: the glass fragments originate from a glass barrier at a shooting range that Suspect 2 has recently been performing construction work on.

This proposition is still competing with $H_p^3$, and $H_p^3$ and $H_d^{3^*}$ are still mutually exclusive. However, they are no longer exhaustive. This is because all possible alternatives to $H_p^3$ are no longer considered by $H_d^{3^*}$. In other words, it is now possible for both $H_p^3$ and $H_d^{3^*}$ to be false. This could happen if the suspect lies or is mistaken about their alternative version of events.

These properties of propositions are important when we begin to consider probabilities associated with them in Chapter [...]. But first we will look at different categories of proposition. 

## Hierarchy of propositions

It's helpful to further categorise propositions based upon the type of assertion that they make. This categorisation makes the proposition's claims explicit and clear, which helps us understand how the proposition can help to answer a question.  

Propositions are mostly grouped into four categories (or levels): offence, activity, source, and sub-source. This is known as the **hierarchy of propositions**, since each category is generally more or less important than the others when considering the case as a whole. Examples of competing propositions for each of these levels are shown in Table \@ref(tab:propositions-table) below.

```{r propositions-df, include=FALSE}
Level <- c(rep("Offence", 4),
           rep("Activity", 4),
           rep("Source", 4),
           rep("Sub-source", 4))
Examples <- c("$H_p^1$: The defendant committed murder",
              "$H_d^1$: The defendant did not commit murder",
              "$H_p^2$: The defendant assaulted the claimant",
              "$H_d^2$: The defendant did not assault the claimant",
              "$H_p^3$: The defendant fired the gun",
              "$H_d^3$: The defendant did not fire the gun",
              "$H_p^4$: The defendant kicked the claimant",
              "$H_d^4$: The defendant was not present when the claimant was kicked",
              "$H_p^5$: The glass from the defendant's clothing originated from the smashed window",
              "$H_d^5$: The glass from the defendant's clothing originated from some other source",
              "$H_p^6$: The blood on the defendant's shoes came from the claimant",
              "$H_d^6$: The blood on the defendant's shoes came from someone else",
              "$H_p^7$: NA",
              "$H_d^7$: NA",
              "$H_p^8$: The DNA on the claimant's clothing came from the defendant",
              "$H_d^8$: The DNA on the claimant's clothing came from someone else")
propsDf <- dplyr::tibble(Level, Examples)
```

```{r propositions-table, echo=FALSE}
options(kableExtra.html.bsTable = T)
propsDf %>%
  knitr::kable(booktabs = TRUE, escape = F, align = "l",
             caption = 'Propositions levels and some examples of competing propositions.') %>%
  kable_styling(c("striped", "condensed"), 
                latex_options = "striped") %>%
  collapse_rows(columns = 1, valign = "top")


```

**Offence level** propositions directly refer to a criminal offence. From Table \@ref(tab:propositions-table) we have:

- $H_p^1$: The defendant committed murder,
- $H_d^1$: The defendant did not commit murder.

We saw these propositions in the case scenario of the previous Section. These competing propositions refer to the **ultimate issue** in the case. Others, such as $H_p^2$ and $H_d^2$ might not be the ultimate issue of the case, but will still determine the fate of the defendant. For this reason, they are the most important level of proposition. 

Offence level propositions are considered by the fact finder in a case and rarely the forensic scientist. This is because propositions about criminal offences fall outside the domain of an scientific expert witness, who gives evidence only on scientific findings within a case. There will be other non-scientific evidence in the case that will need to be considered too. The scientist can indirectly assist in finding the truth of offence level propositions through assisting the court on other related propositions which do fall within their expert domain.

**Activity level** propositions relate to activies and do not assert criminal liability. This leaves them one step below the offence-level propositions in terms of importance to the court; the activity may have been criminal but it also may not have been. In Table \@ref(tab:propositions-table) we have:

- $H_p^4$: The defendant kicked the claimant,
- $H_d^4$: The defendant was not present when the claimant was kicked.

If $H_p^4$ is true, then this still may have been done within the law. However, if $H_p^4$ is proven true to the fact finder, then it is a large step closer to answering the ultimate issue -- all that remains is to show that the activity amounted to assault, thus addressing $H_p^2$. This means that activity level propositions still have a high degree of importance.

Forensic scientists do consider activity level propositions where scientific findings can assist the court on them. In fact, this is the advised level of proposition where it is possible. In order to consider activity level propositions, it is often necessary for the scientist to take in to account contextual information from case circumstances as well as the available scientific findings. If the necessary contextual information is unavailable, then the scientist may find themselves unable to assist the court regarding alleged activities. Consider the propositions:

- $H_p^3$: The defendant fired the gun,
- $H_d^3$: The defendant did not fire the gun. 

Finding gunshot residue particles on the defendant's clothing might not be enough information for the scientist to assist the court on these propositions. They would have to consider other explanations for the defendant having gunshot residue on their clothes first. If the defendant was in the military, part of a shooting club, or had any other reasonable explanation for having this residue on their clothing then the scientist would need to know. They might then require further information about the last time the defendant was in contact with firearms. The amount of particles found on the clothing, and the location in which they were found (e.g. cuffs), might then be considered in light of the defendant's response. Results from the scientific literature about the background abundance and transfer and persistence of gunshot residue particles might then need to be used in addition to the scientific findings from the case. 

[add a picture to the above pls, it is sooo dry]

**Source level** propositions relate to the origin of traces and impressions found, e.g. at a crime scene. The source level makes attributions such as blood to an individual, fibres to a garment of clothing, footwear marks to a piece of footwear, etc. These propositions are relatively simple compared to activity level since they do not comment on questions such as "how?" or "why?" the traces and impressions may have come to be where they were recovered from. The answer to these questions are often the most consequential because they are logically closer to the ultimate issue. Since the activity level addresses these questions and the source level does not, the source level is placed below the activity level in the hierarchy.

Consider the competing propositions from Table \@ref(tab:propositions-table):

- $H_p^5$: The glass from the defendant's clothing originated from the smashed window,
- $H_d^5$: The glass from the defendant's clothing originated from some other source.

These propositions have been derived from evidence that has been found. This is a common situation for source level propositions: trace evidence is found and we need to know its origin. It is also very common for forensic scientists to consider source level propositions. This is because the science is well-equipped to shed light on them. For $H_p^5$ above, the scientist can compare analytic measurements from the recovered glass fragments to other fragments known to come from the smashed window. For $H_d^5$, the scientist will need to compare analytic measurements of the recovered fragments to fragments from other known sources. 

Although the source level propositions are below the activity level in the hierarchy, this is not to say that they are unimportant. In fact, in individual cases, they may be the most important as they might effectively address the most important activity level propositions. For example, if the fact-finder decides that $H_p^5$ is true based on the expert's evidence, then they may not require further evidence to think that the following activity level proposition is true:

- $H_p^{5+}$: the defendant was present when the glass was smashed. 

In other words, the truth of the source level proposition could have a logical domino effect that leads to the fact-finder being sure of the truth of the most important propositions. 

The **sub-source level** is a particular version of the source level. It makes source attributions of more general traces, such as DNA to an individual. The difference between source and sub-source is that the source level states the specific cellular material (e.g. blood) that DNA would be recovered from, whereas sub-source simply states that DNA was found. This is the result of highly sensitive modern measurement technologies. Taking swabs of items of interest at a crime scene can detect very low levels of DNA even though there was no obvious biological material that was swabbed.

This distinction can be key in case circumstances because it leaves open questions of how the DNA came to be there. The sub-source level propositions in Table \@ref(tab:propositions-table) were:

- $H_p^8$: The DNA on the claimant's clothing came from the defendant,
- $H_d^8$: The DNA on the claimant's clothing came from someone else.

DNA found on clothing may be explained by direct contact and indirect transfer, which might be reasonable in any given case. This means that more obstacles are encountered than source level propositions when trying to move to the activity level. If blood has been found on an item then there are far fewer mechanisms of it coming to be there compared to finding DNA alone. 

The hierarchy of propositions and the degree to which the scientist can aid the court in addressing them using science alone is represented in the Figure [...].


It's worth mentioning again that these are conceptual frameworks with which to consider propositions. Individual case circumstances will determine how important each level of proposition is and the degree to which the scientist can consider them using science alone. 

What role does evidence play here? We have seen that source and sub-source level propositions are formed from evidence and that there is a logical relationship between all levels of proposition. In the next section, we will add evidence into the mix.

## Probative value of evidence

The **probative value** of evidence is the degree to which a piece of evidence can discriminate between competing propositions. If a piece of evidence cannot discriminate between propositions then it is not probative for those propositions. But this does not mean it is not probative of any propositions. Consider the following propositions from Table \@ref(tab:propositions-table):

- $H_p^5$: The glass from the defendant's clothing originated from the smashed window,
- $H_d^5$: The glass from the defendant's clothing originated from some other source,

- $H_p^6$: The blood on the defendant's shoes came from the claimant,
- $H_d^6$: The blood on the defendant's shoes came from someone else,

and the following three pieces of evidence:

- $E_1$: glass analysis reveals a "match" between the fragments found on the defendant's clothing and the smashed window,
- $E_2$: DNA analysis of the blood on the defendant's shoes does not match the claimant,
- $E_3$: gun shot residue particles were found on the defendant's clothing.

Clearly evidence $E_1$ is highly probative of $H_p^5$/$H_d^5$. Intuitively, we expect a match if $H_p^5$ is true. We might also come across a match if $H_d^5$ is true. This could occur when the same type of glass has been used for both the smashed window and the other source, but we might reasonably expect this to be less likely than if the glass fragments really did come from the smashed window. One factor which will affect our expectation is how common that type of glass is in other uses, e.g. are all household windows made from that type?

Evidence $E_2$ is not just probative of $H_p^6$/$H_d^6$, but it is almost completely determinative of it. We absolutely do not expect a "no match" if the claimant truly is the source of the blood; $H_p^6$ is incompatible with the evidence. The only situation in which this is not true is if there has been a mistake with handling and analysing the sample. With sufficient evidence gathering and analysing protocols in place, it is assumed that this is not the case unless there is reasonable evidence to suggest otherwise. On the other hand, we absolutely expect "no match" if the claimant is not the source, i.e. if $H_d^6$ is true.

Evidence $E_3$ is not probative of either pair of competing propositions with the lack of contextual information that we have given here. However, it's not hard to imagine that it could be highly probative of the propositions about firing a gun, e.g. $H_p^3$/$H_d^3$, if case circumstances reflected this. 

As soon as forensic evidence has been recovered then it is trivially certain. The uncertainty that does arise comes from assessing the circumstances which may have led to the evidence being left where it was found, or the origin of the trace/impression which was recovered. This means considering the evidence in light of the competing propositions. We did this qualitatively in the discussion above with $E_1$, $E_2$, and $E_3$ and the propositions $H_p^5$/$H_d^5$ and $H_p^6$/$H_d^6$. Since the truth of the propositions is not known by the those investigating, there is uncertainty. And since there is uncertainty, it can be described quantitatively using the language of probability.

Repeating the points above, we can say that the probability of $E_1$, $E_2$, and $E_3$ is 1; since we observed it we know it to be certain. But what we really need is to condition the observed evidence $E_1$, $E_2$, and $E_3$ on the propositions that counsel have put forward $H_p^5$/$H_d^5$ and $H_p^6$/$H_d^6$. This will help the fact finder determine the truth of the propositions. The probabilities for the evidence change when we condition on propositions as they reflect our uncertainty about observing the evidence under these circumstances. Let's revisit our previous qualitative analysis.

What's the probability of $E_1$ assuming that $H_p^5$ is true? It is 1, we would consider it certain to obtain a match if the recovered fragments do come from the smashed window. What's the probability of $E_1$ assuming that $H_d^5$ is true? This is uncertain. We said that this would depend on factors such as how common the glass type of the smashed window was in other sources of glass that the suspect might reasonably have come into contact with. The ability of the evidence $E_1$ to discriminate between $H_p^5$/$H_d^5$ depends on the probability of $E_1$ assuming that $H_d^5$ is true. If this probability is low then it can discriminate well between the propositions because $E_1$ would be far more likely assuming $H_p^5$ were true rather than if $H_d^5$ were true. If the probability is high (near to 1), then it cannot discriminate well between them because the probabilities are of similar magnitude and so the evidence is similarly likely assuming each proposition to be true. In other words, $E_1$ could have high or low (or anywhere inbetween) probative value.

What's the probability of $E_2$ assuming that $H_p^6$ is true? It is 0, with current DNA technology for single donor full profiles, we consider it impossible to obtain a "no match" if the claimant is the origin of the blood. What's the probability of $E_2$ assuming that $H_d^6$ is true? It is 1 or near to 1. To rephrase this, the evidence $E_2$ is only possible if $H_d^6$ were true (ignoring mistakes in handling evidence), and so the evidence has extremely high probative value in favour of $H_d^6$ compared to $H_p^6$.

What's the probability of $E_3$ assuming that $H_p^6$ is true? It is 1. This is because, given the current contextual information, $H_p^6$ has nothing to do with gun shot residue. It is reasonable to assume (although this might have to be proven in a real case) that they are statistically independent and so the probability of $E_3$ assuming that $H_p^6$ is true is the same as the probability of $E_3$ disregarding $H_p^6$, which is 1 because we observed it. Conditioning on an irrelevant proposition tells us nothing that we didn't already know. The same argument can be applied to $H_d^6$. The evidence $E_3$ cannot discriminate between $H_p^6$/$H_d^6$ and so it does not have probative value for these propositions.

The preceding analysis should motivate that propositions are the key elements which define the probative value of evidence. The same piece of evidence can have no probative value for one set of competing propositions but may be highly probative for another. We have shown that using the language of probability, we can quantify the uncertainty surrounding evidence conditional on propositions. The natural question is then, can we quantify probative value? Yes. This is the topic of the next Chapter, the Likelihood Ratio.

## Changing propositions

It goes without saying that since propositions determine the probative value of a piece of evidence, then changes in propositions can affect probative value. This might seem like an obvious statement to some, but it is an important one to highlight because changes in propositions are common and sometimes subtle. The resulting change in probative value might not be as subtle.

As a case develops and investigations continue, new information comes to light and previous information might update. Changes to propositions which are considered by the scientist reflect this, which is why there is always an explicit statement about it that features in written expert witness reports: "these conclusions are based on current information and may change if new information becomes available."

One common example of this is if the defendant changes their version of events or reveals new relevant activities of theirs. The source level glass propositions $H_p^5$/$H_d^5$ might be the default ones considered by the scientist with a lack of other relevant contextual information. If the suspect then states in interview that they have only worn the same clothing recently once before and it was when they were working on a construction site replacing windows, then $H_d^5$ should be revised to reflect this. The proposition might change to:

- $H_d^{5^*}$: The glass from the defendant's clothing originated from the windows on the construction site.

This might be considered a small alteration to the proposition but it could completely diminish the probative value of the glass match $E_1$. Suppose that the glass type of the smashed window at the crime scene is very rare, it is only used for house windows of a very specific style. That means that when compared to all other reasonable sources of glass, this type of glass is found rarely. The probability of $E_1$ assuming $H_d^{5}$ is low in this situation. But, suppose that the defendant was fitting windows of this same style. Now, it would not be a rare occurrence at all. The probability of $E_1$ assuming $H_d^{5^*}$ to be true would be very high, maybe even 1. The evidence has no probative value with the new contextual information and propositions $H_p^{5}$/$H_d^{5^*}$. This represents a change in the reference class, which we saw back in Chapter [...].

[should be explicit about information $I$]

## Probabilities for propositions

In this Chapter we have been very careful to refer to **probabilities for evidence**. We reasoned that probative value is related to probabilities for evidence assuming, in turn, competing propositions to be true and comparing them to each other. We explicitly did not state **probabilities for propositions**, and particularly probabilities for competing propositions in light of the available evidence. 

This is because it is not currently within the role of the expert witness (in the UK) to do so. The expert witness uses their scientific expertise to comment on the likelihood of observing particular pieces of evidence in light of proposed versions of events. It is then within the fact finder's remit to use this information to determine the truth of propositions themselves. In other words, it is up to the fact finder to reason about their own uncertainty for the propositions in light of the evidence they see put before the court. 

If the fact finder's uncertainty is reduced to "beyond reasonable doubt", for what the fact finder decides are the key propositions, then they should return a verdict in favour of the prosecution. If the same uncertainty is not reduced beyond reasonable doubt, then the verdict should be in favour of the defense. Whether the fact finder decides to reason about this uncertainty quantitatively using probability theory or otherwise is up to them.

To hammer home the point, in this book: the **expert** considers **probabilities for evidence**, the **fact finder** considers **probabilities for propositions**. In the discussions following this Chapter, probabilistic statements that we claim come from an expert will only be about observed evidence.

## More information

## Exercises