# Propositions to frame uncertainty {#propositions}

```{r package-load-propositions, include=FALSE}
source("./code/useful-variables.R")
source("./code/helper-functions.R")
# check and load the libraries
library("igraph")
library("ggthemes")
library("kableExtra")
library("tidyverse")
library("plotly")
library("grid")
library("plyr")

```

When applying probability theory in the context of a legal case, it is important that the hypotheses from both sides of counsel are made precise. This is because they will be used to compare against each other as conditioning information in conditional probabilities. Innocent mistakes in proposing imprecise conditioning information can have large effects on the resulting value of the conditional probability. More importantly, such mistakes can lead to incorrect interpretations of conditional probabilities, such as incorrect beliefs about which events may have occurred and which did not.

In this chapter, we will define what is meant by a proposition, how propositions are useful for framing uncertainty about events, and how propositions are combined with probability to determine the probative value of evidence. 

## Facts-in-issue

To motivate the use of propositions, we begin with a discussion about facts-in-issue. Facts-in-issue are the relevant facts of a specific case that are contested by each side of a legal dispute. In criminal trials in the UK, the facts-in-issue are the contested facts between the prosecution's and defence's version of events, otherwise known as their hypotheses.

There is epistemic uncertainty about the facts-in-issue and so relevant evidence about these issues is presented to the court. The fact finder then considers this evidence to determine the truth of the facts-in-issue, and ultimately the outcome of the case. This means that the language and presentation of facts-in-issue must be precise and accurate. They should allow the fact finder to clearly discriminate between the hypotheses of either side of counsel. The expert scientist must also do so impartially without bias in favour of one side or the other. For scientific evidence, the expert does this by formalising the facts-in-issue into what are known as propositions. 

## Propositions

A proposition is an assertion about a factual state of nature. This means that it is capable of being either true or false. Although not always correct, we will speak of propositions as not only **capable of being** either true or false, but **as** being either true or false. 

The truth of propositions does not depend upon any individual's or consensus belief about them (unless the proposition is specifically about their belief). Consider the following:

1. The Earth orbits the Sun.
2. The Earth is flat.

Proposition 1 is true and proposition 2 is false, and we have scientific evidence to prove this. At a certain time in history, however, common belief about those propositions was false: people thought that the Sun orbited the Earth, and that the Earth was flat. It was true that the Earth orbits the Sun both before and after humans had enough evidence to believe it was true. 

Propositions can still evoke epistemic uncertainty; they are true or false but we might not know which. This means we can apply the tools of probability theory to them. Consider another example:

3. Poker Player B is holding an ace in this round.

In this hypothetical poker game, the proposition is true if Player B is holding an ace and it is false if Player B is not holding an ace in this round. If you are Player B then you will know the truth of this and the proposition is trivial. However, if you are not Player B then you will not know the truth of this - you have epistemic uncertainty.

To play this round well, you require a reasonable belief about this proposition (and others) in order to make decisions which account for your uncertainty. Note that in this example you might **never** know the truth of this proposition. Player B might never turn over their cards nor reveal the answer. The proposition is still either true or false regardless of this.

The truth of some propositions may be conditional on other factors. Consider the following:

4. Boris Johnson's Conservative party govern the UK. 

At the time of writing this, the proposition is true. The UK still exists, the Conservative party exist under that same name and are led by Boris Johnson, and his party governs the UK. This proposition will not be true forever, and may not even be true by the time you are reading this. It is useful to be more specific when formulating propositions so that important factors (e.g. time) are made explicit. Doing this makes propositions less ambiguous. Changing the previous proposition to:

5. On 20th May 2020, Boris Johnson's Conservative party govern the UK,

gives a more specific proposition whose truth is now clear to anyone with the knowledge of who governed the UK on that specific date. However, the proposition is now less general since it only refers to one specific day. Whether this reduced generality is worth the increased clarity or not depends on the question at hand. When case information is provided by investigators to forensic scientists, then the scientists have to balance these factors in order to formulate appropriate propositions for the case (if they are not given them). 

## Competing propositions

When formulating propositions from facts-in-issue, the different sides of an adversarial judicial system form different hypotheses about the events which led to the same factual observation.

For example, as part of a case there might be a DNA sample recovered from a crime scene which was found to match the DNA profile taken from a suspect after their arrest. The prosecution might assert that the defendant was at the location in which the DNA sample was recovered. The defence, on the other hand, might accept that this is the defendant's DNA but instead assert that it was indirectly transferred there, e.g. from physical contact with another individual who was provably at the crime scene. Both of these hypotheses would reasonably lead to the defendant's DNA matching the crime scene sample.

Translating these assertions into propositions gives so-called **competing propositions**. Competing propositions typically come in pairs, one put forward by each side of legal counsel. The standard notation that is used for these propositions is $H_p$ and $H_d$ to represent 'prosecution/defence hypothesis'.

## Case scenario: forming propositions

Police were called to the house of a known local drug dealer after reports by neighbours of a suspected gunshot heard from within the property. They arrived at the scene to find a middle-aged male, the known drug dealer, dead in a bedroom with gunshot wounds. A window at the back of the house on the ground floor was found to be smashed, with the glass having fallen towards the inside of the house. As part of the investigation, they found death-threats on the deceased's public social media account from two males. The males are known to police, having previously been involved in drug-related crimes. Both males are arrested at their known addresses, at which time Suspect 1 confesses to having broken in to the house but not having fired the weapon. He refuses to name any accomplices. Suspect 2 refuses to comment, despite glass-covered clothing being found within a plastic bag in a bin at his residence. After further testing, the clothing is found to be covered in gunshot-residue particles. No weapon is recovered.

Here are three facts-in-issue of the many for this case at this preliminary stage. 

Fact-in-issue 1: whether or not Suspect 2 killed the victim with the intent to kill or cause grievous bodily harm

- $H_p^1$: Suspect 2 committed murder,
- $H_d^1$: Suspect 2 did not commit murder.

Fact-in-issue 2: whether or not Suspect 2 fired the gun in question

- $H_p^2$: Suspect 2 fired the gun,
- $H_d^2$: Suspect 2 did not fire the gun.

Fact-in-issue 3: whether or not the glass fragments found on Suspect 2's clothing originated from the smashed window of the deceased's house

- $H_p^3$: The glass fragments originate from the smashed window,
- $H_d^3$: The glass fragments originate from a source other than the smashed window.

In this example, we have extracted the information from the case scenario, formulated potential facts-in-issue, and then created potential competing propositions. The propositions isolate individual issues that appear in the case scenario and make a clear statement about the competing assertions for each issue. Using this process reduces complex case scenarios into individual unambiguous points of focus. This provides an anchor around which the evidence can be assessed.

For case scenarios such as this, some of the information is potentially irrelevant to the experts' task and there is the potential to introduce cognitive bias. Expert's are trained to mitigate this risk and to follow standard operating procedure's to mitigate this risk by design.

## Proposition properties

There are some key properties of these propositions which we also saw when discussing probabilities: mutual exclusivity and exhaustivity. Consider the second pair of competing propositions again:

- $H_p^2$: Suspect 2 fired the gun,
- $H_d^2$: Suspect 2 did not fire the gun.

These propositions are mutually exclusive since when one is true then the other is necessarily false (and vice versa); they cannot both be true. This occurs when two propositions have no logical overlap. Suspect 2 firing the gun and not firing the gun cannot both be true, they do not overlap. However, mutually exclusive propositions can both be false. If the defence's proposition was instead that 'Suspect 1 fired the gun' (exonerating Suspect 2), then this excludes other possibly reasonable scenarios such as there being a third suspect who fired the gun and was not caught. 'Suspect 2 did not fire the gun' covers this alternative scenario.

These propositions are also exhaustive since they cover the entire set of all possibilities for the event in question. This means that one of the competing propositions must be true -- there is no scenario in which both are false. We know that a gun was fired, so either Suspect 2 fired the gun or they did not. The proposition $H_p^2$ covers one specific event and $H_d^2$ covers everything else. When competing propositions are exhaustive and mutually exclusive then they cover all possible events and are logically separable, meaning that one of them must be true and the other is false. This *either one or the other* property makes competing propositions easier to evaluate.

These propositions were guaranteed to be mutually exclusive and exhaustive because of the way they were constructed. First we specified $H_p$ then we **negated** $H_p$ to get $H_d$. The glass propositions provide another example:

- $H_p^3$: The glass fragments originate from the smashed window,
- $H_d^3$: The glass fragments originate from a source other than the smashed window.

If $H_p^3$ is not true, then the glass fragments must originate from a source other than the smashed window, and so we get $H_d^3$. This makes the propositions necessarily mutually exclusive and exhaustive.

The logic of negating propositions is in line with the burden of proof in UK criminal cases. It is for the prosecution to prove beyond reasonable doubt that their assertion of criminal liability is true. The defence, in proving reasonable doubt about the prosecution assertions need only show that the opposite of the prosecution story has reasonable credibility. Negating prosecution propositions is one logical way of doing this.

There are many situations in which negating the prosecution proposition is inappropriate however, e.g. when the defendant has a genuine alternative narrative for events. In this situation, the following assertion might be used:

- $H_d^{3^*}$: the glass fragments originate from a glass barrier at a shooting range that Suspect 2 has recently been performing construction work on.

This proposition is still competing with $H_p^3$, and $H_p^3$ and $H_d^{3^*}$ are still mutually exclusive. However, they are no longer exhaustive. This is because all possible alternatives to $H_p^3$ are no longer considered by $H_d^{3^*}$. It is now possible for both $H_p^3$ and $H_d^{3^*}$ to be false. This could happen if the suspect lies or is mistaken about their alternative version of events.

```{block eval=FALSE, include=isDynamicOutput()}
## Example: exhaustivity and mutual exclusivity
```
``` {r exhaustive-mutually-exclusive-questions, echo=FALSE}
if(isDynamicOutput()){
  knitr::include_url(paste0(QUESTIONS_HOST,"Exhaustive-Mutually-Exclusive-Propositions.html"), height=800)
}
```

## Hierarchy of propositions

Propositions are mostly grouped into four categories (or levels): offence, activity, source, and sub-source. This is known as the **hierarchy of propositions**, since each category is generally more or less important than the others when considering the case as a whole. Categorising propositions into these levels is particularly helpful when interpreting scientific evidence since the value of this evidence heavily depends on the proposition level. Examples of competing propositions for each of these levels are shown in Table \@ref(tab:propositions-table) below.

```{r propositions-df, include=FALSE}
Level <- c(rep("Offence", 4),
           rep("Activity", 4),
           rep("Source", 4),
           rep("Sub-source", 2))
Examples <- c("$H_p^1$: The defendant committed murder",
              "$H_d^1$: The defendant did not commit murder",
              "$H_p^2$: The defendant assaulted the claimant",
              "$H_d^2$: The defendant did not assault the claimant",
              "$H_p^3$: The defendant fired the gun",
              "$H_d^3$: The defendant did not fire the gun",
              "$H_p^4$: The defendant kicked the claimant",
              "$H_d^4$: The defendant was not present when the claimant was kicked",
              "$H_p^5$: The glass from the defendant's clothing originated from the smashed window",
              "$H_d^5$: The glass from the defendant's clothing originated from some other source",
              "$H_p^6$: The blood on the defendant's shoes came from the claimant",
              "$H_d^6$: The blood on the defendant's shoes came from someone else",
              "$H_p^7$: The DNA on the claimant's clothing came from the defendant",
              "$H_d^7$: The DNA on the claimant's clothing came from someone else")
propsDf <- dplyr::tibble(Level, Examples)
```

```{r propositions-table, echo=FALSE}
options(kableExtra.html.bsTable = T)
propsDf %>%
  knitr::kable(booktabs = TRUE, escape = F, align = "l",
             caption = 'Propositions levels and some examples of competing propositions.') %>%
  kable_styling(c("striped", "condensed"), 
                latex_options = "striped") %>%
  collapse_rows(columns = 1, valign = "top")


```

## Offense level propositions

**Offence level** propositions directly refer to a criminal offence. From Table \@ref(tab:propositions-table) we have:

* $H_p^1$: The defendant committed murder,
* $H_d^1$: The defendant did not commit murder.

We saw these propositions in the case scenario. These competing propositions refer to the **ultimate issue** in the case, which is usually the issue for which the fact-finder gives their verdict. Others, such as 

* $H_p^2$: The defendant assaulted the claimant,
* $H_d^2$: The defendant did not assault the claimant,

might not always be the ultimate issue of a case, but they still determine criminal liability. For this reason, offence level propositions are at the top of the hierarchy. 

Offence level propositions are considered by the fact finder in a case and rarely the forensic scientist. This is because propositions about criminal offences fall outside the remit of an expert witness. There will be other non-scientific evidence in the case that will need to be considered too. The expert can indirectly assist the court in finding the truth of offence level propositions by giving evidence related to propositions which fall within their expertise.

## Activity level propositions

**Activity level** propositions relate to activities and do not assert criminal liability. This makes them one step below the offence-level propositions in the hierarchy; the activity may have been criminal but it also may not have been. In Table \@ref(tab:propositions-table) we have:

- $H_p^4$: The defendant kicked the claimant,
- $H_d^4$: The defendant was not present when the claimant was kicked.

If $H_p^4$ is true, then this may have been done lawfully depending on the context. However, if $H_p^4$ is proven true to the fact finder, then it could provide a large step closer to answering the ultimate issue -- all that remains is to show that the activity amounted to assault, which addresses the offence level proposition $H_p^2$.

Forensic scientists do consider activity level propositions where scientific findings can assist the court on them. In fact, this is the advised level of proposition for the scientist to address due to its importance in the hierarchy. In order to consider activity level propositions, it is often necessary for the expert to take in to account contextual information from case circumstances as well as the available scientific findings. If the necessary contextual information is unavailable, then the scientist may find themselves unable to assist the court regarding alleged activities. The same might also be true if there is a lack of scientific understanding relating to a piece of scientific evidence being interpreted in light of asserted activities. To make sense of this, consider the propositions:

- $H_p^3$: The defendant fired the gun,
- $H_d^3$: The defendant did not fire the gun. 

Finding gunshot residue particles on the defendant's clothing might not be enough information for the scientist to assist the court on these propositions. They might require further information about relevant case circumstances in addition to results from the scientific literature about the background abundance and transfer and persistence of gunshot residue particles. 

## Source level propositions

**Source level** propositions relate to the origin of traces and impressions found, e.g. at a crime scene. The source level makes attributions such as blood to an individual, fibres to a garment of clothing, footwear marks to a piece of footwear, etc. These propositions are relatively simple compared to the activity level since they do not comment on questions such as "how?" or "why?" the traces and impressions may have come to be where they were recovered from. The answers to questions of activity are often the more consequential than questions of source because they are logically closer to questions about offences. For this reason, the source level is placed below the activity level in the hierarchy.

Consider the competing propositions from Table \@ref(tab:propositions-table):

- $H_p^5$: The glass from the defendant's clothing originated from the smashed window,
- $H_d^5$: The glass from the defendant's clothing originated from some other source.

These propositions have been derived from evidence that has been found. This is a common situation for source level propositions: trace evidence is found and we need to know its origin. It is also very common for forensic scientists to consider source level propositions. This is because the science is well-equipped to address them. For $H_p^5$ above, the scientist can compare analytical measurements from the recovered glass fragments to other fragments known to come from the smashed window. For $H_d^5$, the scientist will need to compare analytic measurements of the recovered fragments to fragments from other known sources. 

Although source level propositions are below the activity level in the hierarchy, this is not to say that they are unimportant. In fact, in individual cases, they might effectively address the most important activity level propositions. For example, if the fact-finder decides that $H_p^5$ is true based on the expert's evidence, then they may not require further evidence to think that the following activity level proposition is true:

- $H_p^{5+}$: the defendant was present when the glass was smashed. 

In other words, the truth of the source level proposition could have a logical domino effect that leads to the fact-finder being sure of the truth of important activity level propositions. 

## Sub-source level propositions

The **sub-source level** is a particular version of the source level. It makes source attributions of more general traces, such as DNA to an individual. The difference between source and sub-source is that the source level states the specific cellular material (e.g. blood) that DNA is recovered from, whereas sub-source simply states that DNA has been found. This is the result of highly sensitive modern measurement technologies. Taking swabs of items of interest at a crime scene can detect very low levels of DNA even though there was no obvious biological material that was swabbed. The sub-source level is less-informative than the source level because low levels of DNA can easily be transferred between people and surfaces with minimal, and sometimes no, direct contact. 

This distinction between source and sub-source level can be key in case circumstances because it leaves open questions of how the DNA came to be there. The sub-source level propositions in Table \@ref(tab:propositions-table) are:

- $H_p^7$: The DNA on the claimant's clothing came from the defendant,
- $H_d^7$: The DNA on the claimant's clothing came from someone else.

DNA found on clothing may be explained by direct contact and indirect transfer, which might be reasonable in any given case. This means that more obstacles are encountered than source level propositions when trying to move to the activity level. If blood has been found on an item, and so the source level can be addressed, then there are far fewer mechanisms of it coming to be there compared to finding traces of DNA alone. 

<!-- The hierarchy of propositions and the degree to which the scientist can aid the court in addressing them using science alone is represented in the Figure [...]. -->

The hierarchy of propositions is a general conceptual framework which is helpful when considering different propositions. Individual case circumstances will determine how important each level of proposition is and the degree to which the scientist can consider them. In some cases, source level propositions may effectively address the offence level propositions based on the case circumstances and witness testimony.


```{block eval=FALSE, include=isDynamicOutput()}
## Example: proposition levels

Practice identifying the correct proposition level with the examples below.
```
``` {r proposition-level-questions, echo=FALSE}
if(isDynamicOutput()){
  knitr::include_url(paste0(QUESTIONS_HOST,"Propositions-Examples.html"), height=550)
}
```

## Probative value of evidence {#probative-value}

The **probative value** of evidence is the degree to which a piece of evidence can discriminate between competing propositions. When the evidence is scientific, its probative value is determined by the expert and communicated to the court. It is essential that the expert does this using a well-calibrated and validated process (see Section \ref(reliable-probabilities)), so that the determined probative value does not discriminate between propositions with undue bias. The fact finder then decides the truth of the competing propositions based upon this and any other evidence that is relevant to the propositions. 

In order to determine how much the evidence discriminates between competing propositions, the expert considers the competing propositions from counsel and determines how likely the observed evidence would have been assuming each proposition to be true. This logic addresses the questions: how likely would the observations have been if the prosecution's assertions are true? and how likely would the observations have been if the defence's assertions are true? Both of these questions reflect uncertainty in the evidence conditioned on the propositions of the prosecution and defence. This means that they can also be described as conditional probabilities: probabilities of the evidence conditioned on the propositions and any background information (such as agreed facts).

Highly discriminatory evidence, or high probative value, is achieved when there is a large difference between the probabilities under each proposition. If a piece of evidence cannot discriminate between propositions then it is not probative for those propositions, but it may still be probative of other propositions. Little-to-no probative value occurs when there is little to no difference between the conditional probabilities under each proposition. These probabilities should have their calibration validated to ensure that any discrimination is fair and unbiased.

In the examples in the next section, we will use the idea of a match. The word **match** has a specific meaning when referring to a comparison between a questioned sample and a reference sample, such as comparing a finger mark found at a crime scene with a fingerprint taken from a person of interest to an investigation. A match generally means that a selected number of features have been compared between the two samples and that a critical number of those features have been determined as indistinguishable. For example, comparative features of fingerprint examination include skin ridges on the finger such as the ending of a ridge or the splitting of a ridge into two ridges (known as bifurcation), as well as scars and other damage to the skin. The features that are compared and the critical number that determines a match differ between evidence types. The interpretation of a match requires care because it does not necessarily mean that the two samples originate from the same source or that an alleged activity occurred. 

## Example: probative value

Consider the following propositions from Table \@ref(tab:propositions-table):

- $H_p^5$: The glass from the defendant's clothing originated from the smashed window,
- $H_d^5$: The glass from the defendant's clothing originated from some other source,

- $H_p^6$: The blood on the defendant's shoes came from the claimant,
- $H_d^6$: The blood on the defendant's shoes came from someone else.

Suppose there are also the following three pieces of scientific evidence whose probative value is to be determined:

- $E_1$: glass analysis reveals a match between the fragments found on the defendant's clothing and the smashed window,
- $E_2$: DNA analysis of the blood on the defendant's shoes does not match the claimant's DNA,
- $E_3$: gun shot residue particles were found on the defendant's clothing.

Evidence $E_1$ is highly relevant to the source level propositions $H_p^5$/$H_d^5$. We expect a match if $H_p^5$ is true. This belief translates to a probability for $E_1$ conditioned on $H_p^5$ of 1 or very close to 1. There might also reasonably be a match if $H_d^5$ is true. This would occur when the same type of glass has been used for both the smashed window and the other source. One factor which will affect our expectation is how common the type of glass is in other uses, e.g. are all household windows made from that type of glass? The more common the glass, the greater the probability for $E_1$ conditioned on $H_d^5$. The rarer the glass is, the lower the probability for $E_1$ conditioned on $H_d^5$. Rarer glass types result in a greater difference between the conditional probabilities of the evidence $E_1$ under $H_p^5$ compared to $H_d^5$ and therefore have higher probative value. 

Evidence $E_2$ is highly relevant to source level propositions $H_p^6$/$H_d^6$. In this case, $H_p^6$ is incompatible with the evidence: if the claimant truly is the source of the blood then a "no match" is near impossible. The only situation in which this is not true is if there has been a mistake with handling and analysing the sample, so that it becomes contaminated or is switched with another sample. With sufficient evidence-gathering and analysis protocols in place, it is assumed that this is not the case and this is background information. This belief translates into a probability for evidence $E_2$ conditioned on proposition $H_p^6$ of close to 0. On the other hand, we would expect "no match" if the claimant is not the source, i.e. if $H_d^6$ is true. This is a probability for the evidence $E_2$ conditioned on proposition $H_d^6$ of close to 1. Since the conditional probabilities under each source level proposition are so different, this evidence clearly discriminates between them and so it has high probative value.

Evidence $E_3$ does not appear to be relevant to either pair of competing source level propositions with the lack of contextual information given here. Since it is not relevant, then it also has no probative value for those propositions. However, it should not be dismissed outright since there may be other propositions for which it is highly probative. 

This analysis shows that propositions are key in defining the probative value of scientific evidence. The same piece of evidence can have different values for its conditional probabilities depending upon the competing propositions on which it is conditioned. The same piece of evidence can also have no probative value for one set of competing propositions but may be highly probative for another. 

## Changing propositions

Since propositions determine the probative value of a piece of evidence, then changes in propositions can change the probative value. This might seem obvious, but it is important to highlight because changes in propositions are common and sometimes subtle.

New information may be uncovered as part of an ongoing investigation or even after an investigation has been concluded. Propositions used to determine probative value then change according to any updates in information. For this reason, there is usually a caveat statement added to expert witness reports which makes it clear that any of the expert's conclusions may change in light of new information.

One example of this is if the defendant changes their version of events or reveals previously undisclosed information. The source level glass propositions $H_p^5$/$H_d^5$ might be the default ones considered by the scientist with a lack of other relevant contextual information. However, if the suspect then states in an interview that they have only worn the same clothing recently once before and it was when they were working on a construction site replacing windows, then $H_d^5$ should be revised to reflect this. The defence proposition might change to:

- $H_d^{5^*}$: The glass from the defendant's clothing originated from the windows on the construction site.

This might be considered a small alteration to the proposition but it could diminish the probative value of the glass match $E_1$. Suppose that the glass type of the smashed window at the crime scene is very rare, it is only used for house windows of a very specific style. That means that when compared to all other reasonable sources of glass, this type of glass is found rarely. This makes the probability of $E_1$ conditioned on $H_d^{5}$ low. But, if we use the new information above and suppose that the defendant was replacing windows of this same style then a match seems plausible. In other words, the probability of $E_1$ conditioned on $H_d^{5^*}$ could be close to 1. Since we previously reasoned that the probability of $E_1$ conditioned on $H_p^{5}$ is also close to 1, then there is not much difference between the conditional probabilities with this updated proposition. The evidence now has no probative value with the new contextual information and defence proposition. 

Similarly, changing the level in the hierarchy of propositions almost definitely changes the probative value of the evidence. Consider again the related source and activity level competing propositions from Table \@ref(tab:propositions-table):

- $H_p^4$: The defendant kicked the claimant,
- $H_d^4$: The defendant was not present when the claimant was kicked,

- $H_p^6$: The blood on the defendant's shoes came from the claimant,
- $H_d^6$: The blood on the defendant's shoes came from someone else.

Consider the probative value of a match between the claimant's DNA profile and the blood found on the defendant's shoes. The probability of obtaining this match conditioned on $H_p^6$ is much higher than the probability of the match conditioned on $H_d^6$. This makes the evidence highly probative of these competing propositions. However, it is of more interest to the court to determine whether the defendant kicked the claimant or not. This makes it more important to consider the probative value of the evidence with respect to activity level propositions $H_p^4$/$H_d^4$. Depending on the case circumstances, the evidence of a match may be of little value when considering $H_p^4$/$H_d^4$. For example, the defendant may claim to have been absent for the kicking but helped the claimant after noticing them injured in the street. The defendant claims that the blood must have transferred to their shoes whilst they were helping. The defence proposition should then be altered to reflect this new information and the expert might reasonably determine that the probability of observing a match conditional on this new information is close to 1. This would make the probative value of the match low for propositions $H_p^4$/$H_d^4$. Other evidence, such as blood spatter analysis might then be commissioned for its probative value for these propositions instead.

Evidence which is highly probative at the source level may have its value diminished at the activity level. For this reason, it is important to have clear and unambiguous propositions which are addressed at the appropriate proposition level (usually activity). 

## Probabilities for propositions {#prosecutor-fallacy}

In this Chapter we have been careful to refer to **probabilities for evidence** conditioned on propositions. We reasoned that the probative value of evidence is related to the probabilities for evidence assuming, in turn, competing propositions and then comparing the probabilities with each other. We explicitly did not state **probabilities for propositions**, and particularly not probabilities for competing propositions conditioned on evidence. 

This is because it is generally not within the role of the expert witness (in the UK) to comment on probabilities for propositions. The expert witness uses their scientific expertise to comment on the evidence in light of proposed versions of events. It is then within the fact finder's remit to use this information to determine the truth of propositions. In other words, it is up to the fact finder to reason about their own uncertainty for the propositions in light of expert (and other) evidence that is put before the court. Whether the fact finder decides to reason about this uncertainty using probability theory or otherwise is a matter for them.

One well-known error of reasoning regarding conditional probabilities is the **prosecutor's fallacy**. In probability terms, it occurs when the probability of evidence conditioned on a proposition is mistaken for the probability of a proposition conditioned on evidence. Since the conditioning information in these probabilities is reversed, it is also known as transposing the conditional. 

We can give a concrete example of the prosecutor's fallacy by referring back to the disease testing example from Section \@ref(exm-test). Recall that in this example, there was a diagnostic test for a disease with base rate of 1% in a population of 10,000 people. The test had a sensitivity and specificity of 95%. There are multiple possible opportunities for the prosecutor's fallacy to be made here but we choose the one that confuses the following two probabilities: 

1. the probability of a positive test given that the person tested truly does have the disease,
2. the probability of an individual having the disease given that they test positive.

Probability 1 above is the definition of the sensitivity of the test, which is 0.95. In the example, we reasoned probability 2 to be roughly 0.28. In practice, probability 2 is the one that is useful for determining what healthcare actions should be taken, e.g. should the person be admitted to hospital, told to isolate, sent for further testing etc. depending on the disease. Mistaking these two probabilities means that the interpretation of a positive test changes from the correct interpretation of "even after testing positive, an individual is still unlikely to have the disease" to the incorrect interpretation of "almost everyone who tests positive has the disease." This incorrect interpretation highly overestimates the meaning of a positive test and would result in many people who do not have the disease facing treatment as if they did have the disease. 

When forensic evidence is the subject of the interpretation, then one consequence of the prosecutor's fallacy is to misinterpret the probative value of the evidence. When the expert or legal counsel commits the prosecutor's fallacy, then there are the additional consequences of encroaching on the role of the fact finder by stating a probability for a proposition, and also of misleading the court about the value of the evidence.

<!-- Using the terms from Bayes' rule in Section \@ref(bayes), transposing the conditional is equivalent to mistaking the posterior odds $\frac{\text{probability of } H_p \text{ conditioned on } E}{\text{probability of } H_d \text{ conditioned on } E}$ for the likelihood ratio $\frac{\text{probability of } E \text{ conditioned on } H_p}{\text{probability of } E \text{ conditioned on } H_d}$. This is a problem because Bayes' rule states that the posterior odds must be equal to the likelihood ratio multiplied by the prior odds $\frac{\text{probability of } H_p}{\text{probability of } H_d}$. Transposing the conditional in this way assumes that the prior odds are equal to 1, i.e. $\text{probability of } H_p = \text{probability of } H_d$. Whenever the prior odds are not equal to 1, then the prosecutor's fallacy results in mistakenly overstating or understating the posterior odds. When the prosecutor's fallacy is made by someone other than the fact finder, e.g. by legal counsel or the expert, then they are encroaching on the role of the fact finder as well as stating inaccurate posterior odds. -->

<!-- [worked example - edit prior odds and see how much error is induced] -->

<!-- ## More information -->

```{block eval=FALSE, include=isDynamicOutput()}
## Summary: propositions

Use the activity below to create a summary of the key points from this chapter.
```
``` {r propositions-summary-questions, echo=FALSE}
if(isDynamicOutput()){
  knitr::include_url(paste0(QUESTIONS_HOST,"Propositions-Summary.html"), height=800)
}
```

<!-- ## Research study -->

<!-- If you are taking part in our research study, please return to the survey now to answer the questions about this chapter. -->